---
title: "'cpp11armadillo': An 'R' Package to Use the Armadillo C++ Library"
date: "2024-10-14"
abstract:
  "This article introduces 'cpp11armadillo', an R package that integrates the
  highly efficient Armadillo C++ linear algebra library with R through the
  'cpp11' interface. Designed to offer significant performance improvements
  for computationally intensive tasks, 'cpp11armadillo' simplifies the process
  of integrating C++ code into R. This package is particularly suited for R
  users requiring efficient matrix operations, especially in cases where
  vectorization is not possible. Our benchmark demonstrate substantial speed
  gains over native R functions and Rcpp-based setups."
draft: true
author1: Mauricio Vargas Sepúlveda
affiliation11: Department of Political Science
affiliation12: Munk School of Global Affairs and Public Policy
author2: Jonathan Schneider Malamud
affiliation21: Department of Electrical and Computer Engineering, University of Toronto
address: "1 Devonshire Pl"
city: Toronto
state: Ontario
country: Canada
postcode: M5S 3K7
format:
  pdf:
    pdf-engine: xelatex
    template: "elsarticle-template-harv.tex"
    keep-tex: true
    post-process: "move_pdf.sh"
highlighting-macros: true
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(ggplot2)
```

# Introduction

As R continues to grow in popularity for statistical computing and data
analysis, it can create bottlenecks when working with large datasets. The need
to interface with lower-level languages like C++ to bypass these bottlenecks has
led to the development of tools like 'Rcpp' [@eddelbuettel2011], and more
recently, 'cpp11' [@cpp11].

'cpp11armadillo' is a novel package that builds upon the existing Armadillo C++
library [@sanderson2016] to provide high-level access to efficient linear
algebra operations directly from R. The existing 'RcppArmadillo'
[@eddelbuettel2014] already offers this, and 'cpp11armadillo' offers an
alternative with modern C++ features, where we highlight vendoring, that can
improve performance and simplify the its usage.

# Features

'cpp11armadillo' simplifies C++ integration by leveraging the 'cpp11' package
[@cpp11], which is designed to reduce the complexity of including C++ code in R.
Its core features include:

- High-Performance Matrix Operations: By using Armadillo, cpp11armadillo allows
  users to perform matrix multiplications, inversions, and decompositions with
  minimal overhead. These operations are, for some cases, several times faster
  than their R counterparts.
  
- Simplified Syntax: Armadillo offers a user-friendly syntax similar to that of
  'MATLAB' [@sanderson2016], making it accessible to R users without requiring
  an extensive background in C++.

- Modern C++ Support: 'cpp11armadillo' supports C++11 and newer standards,
  enabling efficient memory management, parallel computations, and advanced
  templates, all integrated into an R workflow.

- Integration with R: The package allows direct manipulation of R data
  structures (e.g., vectors, matrices) in C++, providing a seamless workflow
  between the two languages without requiring manual data conversion (e.g., 
  such as exporting data to CSV files and continously reading them in R to
  use the C++ output or vice versa).

# Background

The R programming language was not designed for high-performance computing, it
was created to offer ready-made statistical functions for the end-user, and
its interpreted nature can lead to slow execution times for computationally
intensive tasks [@wickham2019; @r2024]. The same applies to Python and other
high-level languages.

Compiled languages like C++ offer significant performance improvements due to
their direct access to hardware resources and efficient memory management. The
counterpart is that C++ has a steep learning curve and it is less user-friendly
than R. C++ goals are different, it is designed to be fast and efficient, and
it is not focused on statistical computing but on general-purpose programming
[@stroustrup2012].

However, bottlenecks in R can be solved by using vectorized operations instead
of writing loops, which treats data objects as a whole instead of element-wise
[@burns2011]. There are cases where vectorization is not possible or challenging
to implement [@burns2011], and this is where 'cpp11armadillo' comes in and it
offers data structures and functions that are not available in R
that facilitate the implementation of loops and other operations [@r2024;
@sanderson2016].

Even in spite of these difficulties to work with large datasets or
computationally intensive tasks, R is a growing popularity language for
statistical computing and data analysis, and its community has created tools
to integrate it with SQL databases for memory-efficient data access
[@rpostgres]. There are multiple ways to measure a language popularity, and
one of them is the number of R questions in Stack Overflow, which has been
increasing over the years as shown in the following plot [@stackoverflow2024].

```{r stackoverflow, fig.width = 5, fig.height = 3, echo = FALSE, message = FALSE, warning = FALSE, fig.cap = "R Questions in Stack Overflow increase from 2,260 in 2010 to 28,385 in 2023 with a peak of 59,895 in 2020.", fig.pos = "H"}
library(readr)
library(janitor)
library(dplyr)
library(ggplot2)
library(patchwork)
# library(tintin)

stackoverflow <- read_csv("data/stackoverflow.csv") %>%
  clean_names() %>%
  rename(same_yr_pct = r_percentage) %>%
  mutate(
    prev_yr_pct = r_questions / lag(r_questions) * 100,
  ) %>%
  filter(year > 2009, year < 2024)

# clrs <- tintin::tintin_clrs(option = "tintin in the land of the soviets")
clrs <- c("#165976", "#d04e66")

g1 <- ggplot(stackoverflow, aes(x = year, y = r_questions)) +
  geom_line(color = clrs[1]) +
  geom_point(color = clrs[1]) +
  labs(
    title = "R Questions in Stack Overflow",
    x = "Year",
    y = "Number of Questions (N)"
  ) +
  theme_minimal(base_size = 10)

g2 <- ggplot(stackoverflow, aes(x = year, y = prev_yr_pct)) +
  geom_line(color = clrs[2]) +
  geom_point(color = clrs[2]) +
  labs(
    x = "Year",
    y = "Year-over-Year Growth (%)",
    caption = "Data Source: Adapted from Stack Overflow"
  ) +
  theme_minimal(base_size = 10)

g1 + g2

# ggsave("stackoverflow.svg", width = 5, height = 3, dpi = 300)
```

# Modern C++ features

'cpp11' is a modern rewrite of the C++ interface for R, designed to improve
safety, performance, and ease of use. It enforces copy-on-write semantics to
prevent unintended modifications to data, ensuring that changes to objects do
not affect other references. Additionally, 'cpp11' provides safer access to R's
C API, reducing runtime errors in C++ code. It also supports ALTREP objects for
efficient memory management and deferred computations, making it ideal for
handling large datasets. By using UTF-8 strings throughout, 'cpp11' ensures
robust handling of datasets created in different countries where encodings
vary [@cpp11].

Built on C++11 features like smart pointers and lambdas, 'cpp11' offers a more
straightforward and efficient implementation compared to previous bindings. Its
header-only design eliminates ABI compatibility issues, making it easier to
integrate and manage in projects. 'cpp11' also compiles faster, uses less memory,
and grows vectors more efficiently, optimizing performance when dealing with
large amounts of data. These improvements make 'cpp11' a powerful, streamlined
tool for developers who need reliable, high-performance C++ bindings for R
[@cpp11].

Following from 'cpp11', 'cpp11armadillo' leverages these modern C++ features to
provide a seamless interface between R and the Armadillo library, offering
high-performance linear algebra operations with minimal overhead. Besides the
technical aspects, 'cpp11armadillo' offers vendoring, something not available
in 'RcppArmadillo', which can simplify the installation process and reduce
dependency issues, especially when working in environments with restricted
access to the Internet. For instance, the [Niagara Supercomputer](https://docs.scinet.utoronto.ca/index.php/Niagara_Quickstart)
that we use at the University of Toronto has restricted access to the Internet,
and vendoring can simplify the installation process.

Vendoring is a well-known concept in the Go community, and it consists in
copying the dependency code directly into a project's source tree. This approach
ensures that dependencies remain fixed and stable, preventing any external
changes from inadvertently breaking the project. While vendoring offers
stability, it comes with trade-offs. The primary advantage is that updates to
the 'cpp11armadillo' library will not disrupt existing code, and it also copies
'cpp11' C++ headers. However, the drawbacks include an increase in package size
and the loss of automatic updates, meaning that bug fixes and new features will
only be available when manually updated.

In other words, vendoring allows the package creator to provide a
dependency-free package that can be installed in any environment without
requiring the end user to install 'cpp11' nor 'cpp11armadillo'. This approach
makes 'cpp11armadillo' a dependency for the developer but not for the end
user.

# Usage

To use 'cpp11armadillo', users must first install the package from CRAN or
GitHub. The package includes the Armadillo library, no additional
installation is required. The following code shows how to install the package:

```r
install.packages("cpp11armadillo")

# or
remotes::install_github("pachadotdev/cpp11armadillo")
```

Once installed, users can use the provided package template function to create
a new package that uses C++ code with Armadillo. The package template includes
simple examples and all the necessary files to compile the code and install
the new R package. The following code shows how to create a new package:

```r
cpp11armadillo::create_package("cpp11newpackage")
```

The package [vignettes](https://pacha.dev/cpp11armadillo/index.html) cover the
directories organization and the necessary steps to build an R package with
relatively low setup efforts.

It is important to note that 'cpp11armadillo' only work within an R package, and
it is not possible to compile individual C++ scripts on the fly. This design
choice ensures that the R and C++ code is organized following the organization
described in @wickham2023.

# Benchmarks

In order to evaluate the performance of 'cpp11armadillo', we compared it with
native R functions and 'RcppArmadillo'. We focused on a single test consisting
in computing the Balassa Index [@vargassepulveda2020].

The R code for the Balassa Index computation is as follows:

```r
balassa_r <- function(X) {
  B <- t(t(X / rowSums(X)) / (colSums(X) / sum(X)))
  B[B < 1] <- 0
  B[B >= 1] <- 1
  B
}
```

The C++ code for the Balassa Index computation from an R matrix is as
follows:

```cpp
Mat<double> balassa_armadillo(const doubles_matrix<>& x) {
  mat X = as_Mat(x);
  mat B = X.each_col() / sum(X, 1);
  B = B.each_row() / (sum(X, 0) / accu(X));
  B.elem(find(B < 1)).zeros();
  B.elem(find(B >= 1)).ones();
  return B;
}
```

After implementing the C++ computation in 'cpp11armadillo' and 'RcppArmadillo',
we compared the execution time of obtaining the Balassa index year by year for
the 2002-2020 period, which involves 19 matrices with an approximate dimension
of $230 \times 5,100$. After a build time of 4.4s for 'cpp11armadillo' and 9.5s
for 'RcppArmadillo'. The following plot reveals almost identical execution times
for the Armadillo function called from 'cpp11armadillo' and 'RcppArmadillo' with
a marginal advantage for 'cpp11armadillo', and both are significantly faster
than the base R function.

```{r benchmark, fig.width = 5, fig.height = 3, echo = FALSE, message = FALSE, warning = FALSE, fig.cap = "Balassa Index execution time reveals speed gains of around 50% for Armadillo implementations compared to base R.", fig.pos = "H"}
library(bench)

# bench_setup <- readRDS("data/benchmark-setup.rds")
bench_speed <- readRDS("data/benchmark-speed.rds")

bench_speed <- bench_speed %>%
  mutate(
    library = c("Base R", "cpp11armadillo", "RcppArmadillo")
  ) %>%
  select(time)

bench_speed <- tibble(
  library = c(rep("Base R", 100), rep("cpp11armadillo", 100), rep("RcppArmadillo", 100)),
  time = c(unlist(bench_speed$time[1]), unlist(bench_speed$time[2]), unlist(bench_speed$time[3]))
)

# g1 <- bench_setup %>%
#   mutate(
#     median = bench::as_bench_time(median)
#   ) %>%
#   mutate(
#     library = c("cpp11armadillo", "RcppArmadillo")
#   ) %>%
#   ggplot(aes(x = library, y = median)) +
#   geom_col(fill = clrs[1]) +
#   labs(
#     title = "Balassa Index",
#     subtitle = "Build Time",
#     x = "Library",
#     y = "Time (ms)"
#   ) +
#   theme_minimal(base_size = 10)

g2 <- ggplot(bench_speed, aes(x = library, y = time * 1000)) +
  geom_violin(fill = clrs[1]) +
  labs(
    title = "Balassa Index Execution Time",
    x = "Library",
    y = "Time (ms)",
    caption = "Data Source: Own elaboration"
  ) +
  theme_minimal(base_size = 10)

# g1 / g2
g2

# ggsave("benchmark.svg", width = 5, height = 3, dpi = 300)
```

\newpage

The benchmarks were conducted on a ThinkPad X1 Carbon Gen 9 with the
following specifications:

- Processor: Intel Core i7-1185G7 with eight cores
- Memory: 16 GB LPDDR4Xx-4266
- Operating System: Pop!_OS 22.04 based on Ubuntu 22.04
- R Version: 4.4.1
- BLAS Library: OpenBLAS 0.3.20

## Conclusion

'cpp11armadillo' provides a simple and efficient way to integrate C++ code with
R, leveraging the 'cpp11' package and the Armadillo library. It simplifies
the process of writing C++ code for R users, allowing them to focus on the
logic of the algorithm rather than the technical details of the integration. It
can help to solve performance bottlenecks in 'R' code by using the efficient
linear algebra operations provided by Armadillo in cases where vectorization
is challenging. 'RcppArmadillo' is a popular package for this purpose, it
has existed for nearly a decade, 'cpp11armadillo' offers a different approach
with similar performance.
