---
title: "Chapter 3. The Algebra of Least Squares"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{chapter-3}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Load the data and create subsamples

We process the data with the following code:

```{r}
library(econometrics)

experience <- ch3[, 1] - ch3[, 4] - 6
mbf <- (ch3[, 11] == 2) & (ch3[, 12] <= 2) & (ch3[, 2] == 1) & (experience == 12)
sam <- (ch3[, 11] == 4) & (ch3[, 12] == 7) & (ch3[, 2] == 0)
dat1 <- ch3[mbf, ]
dat2 <- ch3[sam, ]

dim(dat1)
dim(dat2)
```

# First regression

The C++ code for the OLS estimator is as follows (see `src/chapter-3.cpp`):

```cpp
Mat<double> ols_(const doubles_matrix<>& y, const doubles_matrix<>& x) {
  Mat<double> Y = as_Mat(y);  // Col<double> Y = as_Col(y); also works
  Mat<double> X = as_Mat(x);

  Mat<double> XtX = X.t() * X;             // X'X
  Mat<double> XtX_inv = inv(XtX);          // (X'X)^(-1)
  Mat<double> beta = XtX_inv * X.t() * Y;  // (X'X)^(-1)(X'Y)

  return beta;
}

[[cpp11::register]]
doubles_matrix<> ols_mat_(const doubles_matrix<>& y, const doubles_matrix<>& x) {
  Mat<double> beta = ols_(y, x);
  return as_doubles_matrix(beta);
}
```

To use the function, it is required to create a model matrix and a response vector:

```{r}
y <- as.matrix(log(dat1[, 5] / (dat1[, 6] * dat1[, 7])))
X <- cbind(1, dat1[, 4])
ols_mat(y, X)
```

# Second regression

```{r}
y <- as.matrix(log(dat2[, 5] / (dat2[, 6] * dat2[, 7])))
experience <- dat2[, 1] - dat2[, 4] - 6
experience2 <- (experience^2) / 100
X <- cbind(1, dat2[, 4], experience, experience2)
ols_mat(y, X)
```

The slope is the same because the residual regression approach is equivalent to
the original regression.

# Create leverage and influence

The influence function can be written in C++ to practice, and it is a good idea
to separate the residuals from the leverage to keep the code modular:

```cpp
Mat<double> yhat_(const doubles_matrix<>& y, const doubles_matrix<>& x) {
  Mat<double> beta = ols_(y, x);
  return as_Mat(x) * beta;
}

double influence_(const doubles_matrix<>& y, const doubles_matrix<>& x) {
  Mat<double> Y = as_Mat(y);
  Mat<double> X = as_Mat(x);

  Mat<double> e = Y - yhat_(y, x);
  Mat<double> XXi = inv(X.t() * X);
  Mat<double> leverage = sum(X % (X * XXi), 1);
  Mat<double> ones = Mat<double>(Y.n_rows, 1, fill::ones);
  Mat<double> d = (leverage % e) / (ones - leverage);

  return as_scalar(max(abs(d)));
}

[[cpp11::register]]
double influence_dbl_(const doubles_matrix<>& y, const doubles_matrix<>& x) {
  return influence_(y, x);
}
```

```{r}
max(abs(influence_dbl(y, X)))
```

# Regression with restricted sample

This is identical to the OLS estimator, but changing the input data:

```{r}
X_r <- X[X[, 2] < 45, ]
y_r <- y[X[, 2] < 45, ]
ols_mat(y_r, X_r)
```
