---
title: "Chapter 4. Least Squares Regression"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{chapter-3}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Load the data and create subsamples

We process the data with the following code:

```{r}
library(econometrics)

experience <- ch4[, 1] - ch4[, 4] - 6
mbf <- (dat[, 11] == 2) & (dat[, 12] <= 2) & (dat[, 2] == 1) & (experience == 12)
sam <- (dat[, 11] == 4) & (dat[, 12] == 7) & (dat[, 2] == 0)
dat1 <- dat[mbf, ]
dat2 <- dat[sam, ]

dim(dat1)
dim(dat2)
```

# Table 4.1

Using the OLS functions from Chapter 3 (see `src/chapter-3.cpp`), we can
compute the standard errors with the following C++ code:

```cpp
Mat<double> stderr_(const doubles_matrix<>& y, const doubles_matrix<>& x) {
  Mat<double> Y = as_Mat(y);
  Mat<double> X = as_Mat(x);

  Mat<double> beta = ols_(Y, X);
  Mat<double> e = Y - (X * beta);
  Mat<double> XXi = inv(X.t() * X);
  Mat<double> leverage = sum(X % (X * XXi), 1);

  double n = Y.n_rows; // using int then returns the wrong "a" value
  double k = X.n_cols;
  double sig2 = as_scalar((e.t() * e) / (n - k));

  Mat<double> ones = Mat<double>(1, k, fill::ones);
  Mat<double> xx = inv(X.t() * X);

  // Heteroskedastic formula
  Mat<double> v0 = xx * sig2;
  Mat<double> s0 = sqrt(v0.diag());

  // White formula
  Mat<double> u1 = X % (e * ones);
  Mat<double> v1 = xx * (u1.t() * u1) * xx;
  Mat<double> s1 = sqrt(v1.diag());

  // HC1 formula
  double a = n / (n - k);
  Mat<double> v1a = a * (xx * (u1.t() * u1) * xx);
  Mat<double> s1a = sqrt(v1a.diag());

  // HC2 formula
  Mat<double> u2 = X % ((e / sqrt(1 - leverage)) * ones);
  Mat<double> v2 = xx * (u2.t() * u2) * xx;
  Mat<double> s2 = sqrt(v2.diag());

  // HC3 formula
  Mat<double> u3 = X % ((e / (1 - leverage)) * ones);
  Mat<double> v3 = xx * (u3.t() * u3) * xx;
  Mat<double> s3 = sqrt(v3.diag());

  // Bind the results
  Mat<double> result = join_rows(s0, s1, s1a, s2);
  result = join_rows(result, s3); // join_rows admits max 4 arguments
  return result.t();
}

[[cpp11::register]]
doubles_matrix<> stderr_dbl_(const doubles_matrix<>& y,
                             const doubles_matrix<>& x) {
  return as_doubles_matrix(stderr_(y, x));
}
```

To use the function, it is required to declare the `ols_` function from
Chapter 3 in the header `00_main.h`:

```cpp
Mat<double> ols_(const Mat<double>& Y, const Mat<double>& X);
```

Now the function can be used in R:

```{r}
y <- as.matrix(log(dat1[, 5] / (dat1[, 6] * dat1[, 7])))
x <- cbind(1, dat1[, 4])

table41 <- stderr_dbl_(y, x)
rownames(table41) <- c("Heteroskedastic", "White", "HC1", "HC2", "HC3")
colnames(table41) <- c("Intercept", "Education")

round(table41, 3)
```

# Table 4.2

Similar to the first section, we need to create subtables:

```{r}
edu12 <- (ch4[, 4] > 11)
dat3 <- ch4[edu12, ]

marriedF <- (dat3[, 12] <= 3) & (dat3[, 2] == 1)
marriedM <- (dat3[, 12] <= 3) & (dat3[, 2] == 0)
unionF <- (dat3[, 8] == 1) & (dat3[, 2] == 1)
unionM <- (dat3[, 8] == 1) & (dat3[, 2] == 0)
fmarriedF <- (dat3[, 12] <= 6) & (dat3[, 12] > 3) & (dat3[, 2] == 1)
fmarriedM <- (dat3[, 12] <= 6) & (dat3[, 12] > 3) & (dat3[, 2] == 0)
black <- (dat3[, 11] == 2)
american_indian <- (dat3[, 11] == 3)
asian <- (dat3[, 11] == 4)
mixed <- (dat3[, 11] >= 6)
```

The C++ function for Table 4.2 is similar to the previous one:

```cpp
Mat<double> beta_stderr_(const doubles_matrix<>& y, const doubles_matrix<>& x) {
  Mat<double> Y = as_Mat(y);
  Mat<double> X = as_Mat(x);

  Mat<double> beta = ols_(Y, X);
  Mat<double> e = Y - (X * beta);
  Mat<double> XXi = inv(X.t() * X);
  Mat<double> leverage = sum(X % (X * XXi), 1);

  double n = Y.n_rows;  // using int then returns the wrong "a" value
  double k = X.n_cols;

  Mat<double> ones = Mat<double>(1, k, fill::ones);
  Mat<double> xx = inv(X.t() * X);

  // HC2 formula
  Mat<double> u2 = X % ((e / sqrt(1 - leverage)) * ones);
  Mat<double> v2 = xx * (u2.t() * u2) * xx;
  Mat<double> s2 = sqrt(v2.diag());

  // Bind the results
  Mat<double> result = join_rows(beta, s2);
  return result;
}

[[cpp11::register]]
doubles_matrix<> beta_stderr_dbl_(const doubles_matrix<>& y,
                                  const doubles_matrix<>& x) {
  return as_doubles_matrix(beta_stderr_(y, x));
}
```

Now we can replicate the results in R:

```{r}
y <- as.matrix(log(dat3[, 5] / (dat3[, 6] * dat3[, 7])))
experience <- dat3[, 1] - dat3[, 4] - 6
exp2 <- (experience^2) / 100
x <- cbind(
  1, dat3[, 4], experience, exp2, dat3[, 2],
  unionF, unionM, marriedF, marriedM, fmarriedF, fmarriedM,
  dat3[, 3], black, american_indian, asian, mixed
)

table42 <- beta_stderr_dbl_(y, x)
rownames(table42) <- c("Intercept", "Education", "Experience",
  "Experience^2 / 100", "Female", "Female union member", "Male union member",
  "Married female", "Married male", "Formerly married female",
  "Formerly married male", "Hispanic", "Black", "American Indian", "Asian",
  "Mixed race"
)
colnames(table42) <- c("Estimate", "Std. Error")

round(table42, 3)
```

# DDK (2011)

The C++ code for the DDK (2011) function is as follows:

```cpp
Mat<double> ddk_(const doubles_matrix<>& y,
                      const doubles_matrix<>& x,
                      const doubles_matrix<>& z) {
  Mat<double> Y = as_Mat(y);
  Mat<double> X = as_Mat(x);
  Mat<double> Z = as_Mat(z);

  int n = Y.n_rows;
  int k = X.n_cols;

  Mat<double> xx = X.t() * X;
  Mat<double> xx_inv = inv(xx);
  Mat<double> beta = ols_(Y, X);
  
  Mat<double> xe(n, k);
  for (int j = 0; j < k; j++) {
    xe.col(j) = X.col(j) % (Y - X * beta);
  }

  // Unique groups in Z
  Mat<double> Z_unique = unique(Z);
  int n_groups = Z_unique.n_rows;

  // Grouped sums
  Mat<double> xe_sum(n_groups, k, fill::zeros);
  for (int g = 0; g < n_groups; g++) {
    int group = Z_unique(g, 0);
    for (int i = 0; i < n; ++i) {
      if (Z(i, 0) == group) {
        xe_sum.row(g) += xe.row(i);
      }
    }
  }

  Mat<double> omega = xe_sum.t() * xe_sum;
  double scale = (n_groups / (n_groups - 1)) * ((n - 1) / (n - k));
  Mat<double> V_clustered = scale * (xx_inv * omega * xx_inv);
  Mat<double> se_clustered = sqrt(V_clustered.diag());

  return join_rows(beta, se_clustered);
}

[[cpp11::register]] doubles_matrix<> ddk_dbl_(const doubles_matrix<>& y,
                                              const doubles_matrix<>& x,
                                              const doubles_matrix<>& z) {
  return as_doubles_matrix(ddk_(y, x, z));
}
```

Now we can use the function in R:

```{r}
y <- scale(as.matrix(ch4_2$totalscore))
x <- cbind(1, as.matrix(ch4_2$tracking))
z <- as.matrix(ch4_2$schoolid)

ddk <- ddk_dbl_(y, x, z)
rownames(ddk) <- c("Intercept", "Tracking")
colnames(ddk) <- c("Estimate", "Std. Error")

round(ddk, 3)
```
